---
author: Assignment, Dr. Sara Hermann. Code, Jared Adam
execute: 
  echo: false
---

# Testing your skills: ggplot 
## Recreate these plots to the best of your ability {.unnumbered}

In this section, we will be using the **mtcars** data set. 
Let's start by reading in **ggplot** and the **mtcars** data set. 

```{r}
#| echo: true
#| output: false
library(ggplot2)
data("mtcars")
```

For more information about this data set, look in the *help* section. 
```{r}
#| echo: true
#| output: false
?mtcars()
```

Now, let's take a look a little closer at the data set. 
Which 'class' do these variables belong to? 
Are they numeric? Characters? Integers? 

This knowledge is very impoartant when building your plots.

```{r}
#| echo: true
#| output: false

View(mtcars)
str(mtcars)
```

### Hints {.unnumbered}

Here is an example of how to change the variable type inside of the ggplot lines **and** outside. 
Notice here, we are calling the *am* column with the **$** operator. 

```{r}
#| output: false
#| echo: true
ggplot(mtcars, aes(x = as.factor(am)))

mtcars$am <- as.factor(mtcars$am)
```

You may also need to search for ggplot format help on these. 
Please see the extra resources at the bottom of this page. 

### Plot time {.unnumbered}

#### Graph 1: geom_point {.unnumbered}

```{r}
#| warning: false
#| 

ggplot(data=mtcars, aes(x=wt, y=mpg))+
  geom_point(color="blue", shape=15) + #changing data to blue squares with shape=15
  stat_smooth(method=lm, se=TRUE, color="black")+ #changing line to black 
  labs(x = "Weight (1000 lbs)",#axis titles 
  y = "Miles/(US) gallon")+
  theme(axis.text = element_text(size=8), #changing the font sizes of axis
        axis.title.x = element_text(size=10),#changing the font sizes of axis titles
        axis.title.y = element_text(size=10))
```

#### Graph 2: geom_boxplot {.unnumbered}

```{r}
#| warning: false
mtcars$am = as.factor(mtcars$am)
is.factor = mtcars$am
levels(mtcars$am)
am_labels = factor(mtcars$am, labels=c("Automatic", "Manual"))

ggplot(data=mtcars, aes(x=disp, y=am_labels, fill=factor(gear)))+
  geom_boxplot(fatten=2.5)+ #making the median line wider (fatter) because the one in the picture looks bigger
  theme( 
  panel.grid.major = element_line(size = 0.5, linetype = "solid", color = "grey"),#major lines = grey 
  panel.grid.minor = element_line(size = 0.5, linetype = "solid", color = "grey"),#minor lines = grey 
  panel.background = element_rect(fill = "white", color = "black", size = 0.5, linetype = 'solid'))+ #backgroun =white, border = black 
  scale_x_continuous(name="Displacement (cu.in.)",limits = c(50,max(mtcars$disp)))+ #axis names and legend label
  scale_y_discrete(name="Transmission")+
  labs(fill="Number of forward gears")
```  


#### More to come!! 6/9/2023 {.unnumbered}

### Additional resources {.unnumbered}

[Lecture notes](https://www.jeffdoser.com/files/for875/_book/ggplot2.html)

[Basic R graphics guide](https://www.jeffdoser.com/files/for875/_book/graphics.html)

[ggplot online textbook](https://github.com/hadley/ggplot2-book)